---
title: "Regression Analysis and Diagnostic Tests"
author: "Lumumba W. Victor"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      fig.height = 6, fig.width = 8)
```

# **Regression and Correlational Analysis in R**

## **Regression Analysis**

Regression analysis is a statistical technique used to examine the
relationship between one or more independent variables (predictors) and
a dependent variable (outcome). It helps to understand how changes in
the independent variables are associated with changes in the dependent
variable. Regression analysis can be used for prediction, forecasting,
and identifying important factors that influence the outcome.

## **History of Regression Analysis**

Regression analysis has a long history dating back to the 19th century.
The term "regression" was first introduced by Sir Francis Galton in the
late 1800s when he observed that the heights of children tended to
regress toward the mean height of their parents. This concept laid the
foundation for the development of regression analysis as a statistical
method.

Karl Pearson and Ronald Fisher further contributed to the development of
regression techniques in the early 20th century, leading to the
widespread use of regression analysis in various fields such as
economics, social sciences, and natural sciences.

## **Correlation Analysis**

Correlation analysis, on the other hand, is a statistical method used to
measure the strength and direction of the relationship between two
variables. It quantifies the degree to which changes in one variable are
associated with changes in another variable. Correlation analysis does
not imply causation; it only indicates the presence and strength of a
relationship.

## **Distinguishing Correlation and Causation**

The two concepts of correlation and causation are often confused, but
they are distinct. Correlation indicates a statistical association
between two variables, while causation implies that changes in one
variable directly cause changes in another. It is important to note that
correlation does not imply causation, and further research is often
needed to establish causal relationships.

## **Set up the Current Working Directory**

```{r}
setwd("D:/Research Workshop and Seminars and Trainings/MEDIACREST COLLEGE/Advanced Data Analysis and AI-August Intake/Week-2-Sessions")
```

### **Explore the Data set**

```{r}
# Load data
data(mtcars)
head(mtcars)
```

```{r}
# View structure
str(mtcars)
```

### **Description of the Dataset**

mtcars {datasets} R Documentation Motor Trend Car Road Tests

**Description**

The data was extracted from the 1974 Motor Trend US magazine, and
comprises fuel consumption and 10 aspects of automobile design and
performance for 32 automobiles (1973–74 models).

Usage mtcars Format A data frame with 32 observations on 11 (numeric)
variables.

$$, 1$$ mpg Miles/(US) gallon $$, 2$$ cyl Number of cylinders $$, 3$$
disp Displacement (cu.in.) $$, 4$$ hp Gross horsepower $$, 5$$ drat Rear
axle ratio $$, 6$$ wt Weight (1000 lbs) $$, 7$$ qsec 1/4 mile time $$,
8$$ vs Engine (0 = V-shaped, 1 = straight) $$, 9$$ am Transmission (0 =
automatic, 1 = manual) $$,10$$ gear Number of forward gears $$,11$$ carb
Number of carburetors

## **Scatter Plots are Essential Before Regression and Correlation**

```{r}
library ( ggplot2 )

# Scatterplot : Weight vs MPG
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point( color ="blue", size =1) +
  geom_smooth( method ="lm", se=TRUE, color ="red") +
  labs (title =" Relationship between Weight and MPG",
         x=" Weight (1000 lbs)", y=" Miles per Gallon ")
```

\*\* NOte\*\* There is a strong negative correlation between weight and
miles per gallon. As the weight of the car increases, the miles per
gallon decreases.

## **Comput Pearson Correlation Analysis in R**

```{r}
# Compute Pearson correlation
cor(mtcars$wt, mtcars$mpg, method = "pearson")
```

### **Test significance**

### **State the null and alternative hypotheses:**

- H0: There is no significant correlation between weight and miles per
  gallon (r = 0).
- H1: There is a significant correlation between weight and miles per
  gallon (r ≠ 0).

### **Decision Rule:**

- If the p-value is less than the significance level (commonly set at
  0.05), we reject the null hypothesis. Otherwise, we fail to reject the
  null hypothesis.

### **Testing the Significance of the Correlation**

```{r}
cor.test(mtcars$wt, mtcars$mpg, method = "pearson")
```

**Interpretation:** The Pearson correlation coefficient (r) is -0.867,
indicating a strong negative correlation between weight and miles per
gallon. The p-value is less than 0.05, which means we reject the null
hypothesis and conclude that there is a statistically significant
correlation between weight and miles per gallon.

### **When Normality is Violated or Data is Ordinal**

```{r}
# Compute Spearman correlation
cor(mtcars$wt, mtcars $mpg, method = "spearman")
```

```{r}
# Test significance
cor.test(mtcars$wt, mtcars$mpg, method = "spearman")
```

### **Correlation Metrix**

### **Examine relationships between multiple variables simultaneously.**

```{r}
# Select numeric columns
num_vars <- mtcars[, c("mpg", "cyl", "disp", "hp", "wt")]

# Correlation matrix
cor_matrix <- cor(num_vars, method = "pearson")
print(round(cor_matrix, 2))
```

```{r}
# Select numeric columns
num_vars <- mtcars[, c("mpg", "cyl", "disp", "hp", "wt")]

# Correlation matrix
cov_matrix <- cov(num_vars, method = "pearson")
print(round(cov_matrix, 2))
```

### **Visualize Correlation Matrix**

```{r}
library(corrplot)
corrplot(cor_matrix, 
         #title = "Pairwise Correlation Plot",
         method = "color", 
         #type = "upper",
         addCoef.col = "white",
         tl.col = "black", 
         tl.srt = 45)
```

### **SLR Model in R**

```{r}
## Fit Simple Linear Regression
model_slr <- lm(mpg ~ wt, data = mtcars )
```

```{r}
summary(model_slr)
```

### **State the Null and Alternative Hypotheses:**

- H0: There is no statistically significant effect of weight of the car
  on miles per gallon (the coefficient for weight is equal to zero) at
  5% level of significance.

- H1: There is a statistically significant effect of weight of the car
  on miles per gallon (the coefficient for weight is not equal to zero)
  at 5% level of significance.

### **Simple Linear Regression Equation**

$$\text{Equation:} = \hat{MPG} = 37.2851 -5.3445 \times \text{wt}$$

**Interpretation:** The coefficient for weight (wt) is -5.344, which
means that for every 1 unit increase in weight (1000 lbs), the miles per
gallon (mpg) decreases by approximately 5.344 units, holding all else
constant. The p-value for the weight variable is less than 0.05,
indicating that the effect of weight on miles per gallon is
statistically significant at 5% level of significance.

### **Null and Alternative Hypotheses for the Overall Model (F-Test):**

- H0: The overall regression model is not statistically significant (the
  model does not explain a significant portion of the variance in miles
  per gallon).

- H1: The overall regression model is statistically significant (the
  model explains a significant portion of the variance in miles per
  gallon).

### **Decision Rule:**

- If the p-value for the F-test is less than the significance level
  (commonly set at 0.05), we reject the null hypothesis. Otherwise, we
  fail to reject the null hypothesis.

**Interpretation of F-Test:** The F-statistic is 91.38 with a p-value of
1.294e-10, which is less than 0.05. Therefore, we reject the null
hypothesis and conclude that the overall regression model is
statistically significant, meaning that weight of the car significantly
explains the variance in miles per gallon.

### **Report R-squared and Adjusted R-squared:**

- R-squared: 0.7528
- Adjusted R-squared: 0.7446 **Interpretation:**

The R-squared value of 0.7528 indicates that approximately 75.28% of the
variance in miles per gallon (mpg) can be explained by the weight of the
car (wt). The adjusted R-squared value of 0.7446 accounts for the number
of predictors in the model and provides a more accurate measure of the
model's explanatory power. In this case the adjusted R-squared is
slightly lower than the R-squared, which suggests that the model is a
good fit for the data, but there may be some room for improvement by
including additional predictors.

### **Confidence Intervals for the Coefficients:**

```{r}
confint(model_slr, level = 0.95)
```

**Interpretation:** The 95% confidence interval for the intercept
(37.2851) ranges from 33.45 to 41.12, which means we are 95% confident
that the true intercept value lies within this range. The 95% confidence
interval for the weight coefficient (-5.3445) ranges from -6.49 to
-4.20, which means we are 95% confident that the true effect of weight
on miles per gallon lies within this range. Since the confidence
interval for the weight coefficient does not include zero, it further
supports the conclusion that weight has a statistically significant
effect on miles per gallon at the 5% level of significance.

### **Prediction Using the SLR Model:**

```{r}
# New data frame
new_car <- data.frame(wt = 3.5)
new_car
```

### **Narrow confidence interval for the mean response**

```{r}
# Predict with Confidence Interval
predict(model_slr, newdata = new_car, 
         interval = "confidence")
```

#### **Wider confidence interval for the mean response**

```{r}
# Predict with Prediction Interval ( wider )
predict(model_slr, newdata = new_car ,
        interval = "prediction")
```

## **Using Hypertension Data Set to Develop SLR**

```{r}
# Load data
htn <- read.csv("htn_dat.csv")
head(htn)
```

### \*\*Explore the Data

```{r}
## Check the Structure of the Data
str(htn)
```

### \*\*Check the Presence of Missing Value

```{r}
# Check for missing values as a dataframe
Missing_Values <- data.frame(Missing.values = colSums(is.na(htn)))
Missing_Values
```

### **Alternatively**

```{r}
# Check for missing values as a dataframe
missing_values <- data.frame(
  Variable = names(htn),
  Missing_Count = sapply(htn, function(x) sum(is.na(x))),
  Missing_Percentage = sapply(htn, function(x) mean(is.na(x)) * 100)
)
missing_values
```

### **Check Distribution of BMI**

```{r}
ggplot(htn, aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of BMI", x = "BMI", y = "Frequency")
```

### **Summmary Statistics to Check Skeweness**

```{r}
library(psych)
describe(htn)
```

### **Handle Missing Values**

i.  Approach 1: Remove rows with missing values (complete case analysis)

```{r}
comp_data <- na.omit(htn)
head(comp_data)
```

### **Check Missing Values after Cleaning**

```{r}
data.frame(colSums(is.na(comp_data)))
```

### **Scatter Plot to Show the Relationship between SBP and BMI**

```{r}
ggplot(comp_data, aes(x = BMI, y = SBP)) +
  geom_point(color = "steelblue", size = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Relationship between BMI and SBP",
       x = "Body Mass Index", y = "Systolic Blood Pressure")
```

### Estimate the SLR Model

```{r}
model_htn <- lm(SBP ~ BMI, data = comp_data, level = 0.95)
summary(model_htn)
```

### **The Equation of the SLR Model**

$$\hat{SBP} = 92.84148 + 0.85717 \times BMI$$

### **Ranges for BMI**

- Normal BMI: 18.5 to 24.9
- Overweight BMI: 25 to 29.9
- Obese BMI: 30 or higher

### **The BMI of the New Patient**

```{r}
bmi= (75)/(1.75^2)
bmi
```

### **Predict the BMI of the Person above**

```{r}
new_person <- data.frame(BMI = bmi)
new_person
```

```{r}
# Predict SBP with Confidence Interval
predict(model_htn, newdata = new_person, interval = "confidence")
```

```{r}
# Predict SBP with Prediction Interval
predict(model_htn, newdata = new_person, interval = "prediction")
```

### **Multiple Linear Regression (MLR) Model**

- Predictors: wt and hp
- Outcome: mpg

### **State the Null and Alternative Hypotheses for the MLR Model:**

- H0: There is no statistically significant effect of weight and
  horsepower on miles per gallon (the coefficients for weight and
  horsepower are equal to zero) at 5% level of significance.

- H1: There is a statistically significant effect of weight and
  horsepower on miles per gallon (the coefficients for weight and
  horsepower are not equal to zero) at 5% level of significance.

### **Select of Level of Significance**

- Commonly set at 0.05

#### **Collect the Data**

#### **Run the Test (Multple Linear Regression Model)**

```{r}
model_mlr <- lm(mpg ~ wt + hp, data = mtcars)
summary(model_mlr)
```

### **The Equation of the MLR Model**

$$\hat{MPG} = 37.22727 - 3.877 \times \text{wt} - 0.031 \times \text{hp}$$

### **Decision Rule for the MLR Model:**

> If the p-value for a predictor is less than the significance level
> (commonly set at 0.05), we reject the null hypothesis for that
> predictor. Otherwise, we fail to reject the null hypothesis for that
> predictor.

### **Interpretation of the MLR Model:**

- The coefficient for weight (wt) is -3.877, which means that for every
  1 unit increase in weight (1000 lbs), the miles per gallon (mpg)
  decreases by approximately 3.877 units, holding horsepower (hp)
  constant. The p-value for weight is less than 0.05, indicating that
  the effect of weight on miles per gallon is statistically significant
  at 5% level of significance.

- The coefficient for horsepower (hp) is -0.031, which means that for
  every 1 unit increase in horsepower, the miles per gallon (mpg)
  decreases by approximately 0.031 units, holding weight (wt) constant.
  The p-value for horsepower is less than 0.05, indicating that the
  effect of horsepower on miles per gallon is statistically significant
  at 5% level of significance.

### **Reporyt R-squared and Adjusted R-squared for the MLR Model:**

- An R-squared value of 0.8268 indicates that approximately 82.68% of
  the variance in miles per gallon (mpg) can be explained by the weight
  and horsepower of the car. The adjusted R-squared value of 0.8148
  accounts for the number of predictors in the model and provides a more
  accurate measure of the model's explanatory power. In this case, the
  adjusted R-squared is slightly lower than the R-squared, which
  suggests that the model is a good fit for the data, but there may be
  some room for improvement by including additional predictors or
  considering interactions between predictors.

- An adjusted R-squared value of 0.8148 indicates that approximately
  81.48% of the variance in miles per gallon (mpg) can be explained by
  the weight and horsepower of the car, after adjusting for the number
  of predictors in the model. This suggests that the model has a strong
  explanatory power, but there may still be some unexplained variance
  that could potentially be accounted for by including additional
  predictors or considering interactions between predictors.

## **Model Diagnostics for the MLR Model**

> Multicollinearity

Multicollinearity occurs when two or more predictor variables in a
multiple regression model are highly correlated with each other. This
can lead to unstable estimates of the regression coefficients and make
it difficult to determine the individual effect of each predictor on the
outcome variable.

```{r}
library (car)
# Calculate VIF
vif(model_mlr)
```

### **Report VIF Values:**

- VIF values for weight (wt) and horsepower is 1.766625 indicating that
  there is no significant multicollinearity between the predictors in
  the model. Generally, a VIF value greater than 5 or 10 is considered
  indicative of high multicollinearity, so in this case, the VIF values
  suggest that multicollinearity is not a concern for this model.

### **Diagnostic Plots for the MLR Model**

```{r, fig.height=8, fig.width=10}
# Generate diagnostic plots
par(mfrow =c(2 ,2))
plot(model_mlr)
par(mfrow =c(1 ,1))
```

### **Identify the Influential Observation as Indicared by Cook's Distance**

```{r}
# Calculate Cook's distance
cooks_distance <- cooks.distance(model_mlr)
# Identify influential observations
influential_obs <- which(cooks_distance > (4 / nrow(mtcars)))
influential_obs
```

### **Remove Influential Observation and Refit the Model**

```{r}
# Remove influential observation
mtcars_no_influential <- mtcars[-influential_obs, ]
# Refit the model without influential observation
model_mlr_no_influential <- lm(mpg ~ wt + hp, data = mtcars_no_influential)
summary(model_mlr_no_influential)
```

### **Inferential Test for Normality and Homoscedasticity**

#### **Null and Alternative Hypotheses for Normality:**

- H0: The residuals of the model are normally distributed.
- H1: The residuals of the model are not normally distributed.

```{r}
# Normality test (Shapiro-Wilk test)
shapiro.test(residuals(model_mlr))
```

```{r}
# Normality test (Anderson-Darling test)
library(nortest)
ad.test(residuals(model_mlr))
```

```{r}
ggplot(data = data.frame(residuals = residuals(model_mlr)), aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot of Residuals", 
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles")
```

```{r}
ggplot(data = data.frame(residuals = residuals(model_mlr)), aes(x = residuals)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  labs(title = "Histogram of Residuals", 
       x = "Residuals", 
       y = "Frequency")
```

#### **Null and Alternative Hypotheses for Homoscedasticity:**

- H0: The residuals of the model have constant variance
  (homoscedasticity).
- H1: The residuals of the model do not have constant variance
  (heteroscedasticity).

```{r}
# Homoscedasticity test (Breusch-Pagan test)
library(lmtest)
bptest(model_mlr)
```

#### **Null and Alternative Hypotheses for Independence:**

- H0: The residuals of the model are independent (no autocorrelation).
- H1: The residuals of the model are not independent (autocorrelation
  present).

```{r}
# independence test (Durbin-Watson test)
dwtest(model_mlr)
```

## **MULTIPLE LINEAR REGRESSION MODEL**

### **US Macroeconomic Variables**

```{r}
library(ggplot2)
data("economics")
head(economics)
tail(economics)
```

```{r}
?economics
```

### **US economic time series**

#### **Description**

This dataset was produced from US economic time series data available
from <https://fred.stlouisfed.org/>. economics is in "wide" format,
economics_long is in "long" format.

Usage economics

economics_long Format A data frame with 574 rows and 6 variables:

date Month of data collection

pce personal consumption expenditures, in billions of dollars,
<https://fred.stlouisfed.org/series/PCE>

pop total population, in thousands,
<https://fred.stlouisfed.org/series/POP>

psavert personal savings rate,
<https://fred.stlouisfed.org/series/PSAVERT/>

uempmed median duration of unemployment, in weeks,
<https://fred.stlouisfed.org/series/UEMPMED>

unemploy number of unemployed in thousands,
<https://fred.stlouisfed.org/series/UNEMPLOY>

An object of class tbl_df (inherits from tbl, data.frame) with 2870 rows
and 4 columns.

### **Estimate the Multiple Linear Regression Model**

```{r}
options(scipen = 999)
model_mlr <- lm(psavert ~ pce + 
                  unemploy + 
                  pop + 
                  uempmed, 
                data = economics)
summary(model_mlr)
```

### **Hypothesis Testing for Overall Model**

> Null hypothesis: The null model(intercept only model) is a better fit
> Alternative hypothesis: The full fitted model is a better fit

$$\hat{\text{Psavert}} = 45.67 + 0.00085(\text{pce}) + 
0.00024(\text{umploy}) - 0.00017(\text{pop}) + 0.166(\text{uempmed})$$

### **Report the Results**

```{r}
library(report)
report(model_mlr)
```

### **Marginal Effect of the Linear Effect**

```{r}
model_mlr2 <- lm(log(psavert) ~ log(pce) + 
                  log(unemploy) + 
                  log(pop) + 
                  log(uempmed), 
                data = economics)
summary(model_mlr2)
```

### **Model Diagnostics**

```{r}
model_new <- lm(log(psavert) ~ log(pce) + 
                  log(unemploy) + 
                  log(uempmed), 
                data = economics)
summary(model_new)
```

### **Normality Test for the Residuals**

> Null hypothesis: The residuals follows a normal distribution
> Alternative hypothesis: The residuals deviates from a normal
> distribution

```{r}
resid <- residuals(model_new)
shap_test <- shapiro.test(resid)
shap_test
```

```{r}
resid_df <- data.frame(Residuals = resid)
head(resid_df,5)
```

```{r}
ggplot(data = resid_df, aes(x = Residuals))+
  geom_histogram(fill = "steelblue",
                 color = "black",
                 bins = 25,
                 alpha = 0.9)+
  labs(
    title = "Histogram Showing the Distribution of Residuals",
    x = "Residuals",
    y = "Frequency"
  )
```

### **Heteroscedasticity**

> Null: Residuals have a constant variance over time Alternative: The
> variance of the residuals is not constant over time

```{r}
# Homoscedasticity test (Breusch-Pagan test)
library(lmtest)
bptest(model_new)
```

### **Multicollinearity**

```{r}
library(car)
vif(model_new)
```

## **Additional Example: Macroeconomic Variables**

```{r}
df = read.csv("training model.csv")
head(df, 10)
```

### **Convert year variable to actual date format**

```{r}
df$year <- as.Date(df$year)
head(df,10)
```

### **Run the Multiple Linear Model**

```{r}
model <- lm(CPI ~ ExchRate+LendIntRates, data = df)
summary(model)
```

```{r}
library(report)
report_statistics(model)
```

```{r}
report(model)
```

## **Model Diagnostics for the MLR Model**

> Multicollinearity

Multicollinearity occurs when two or more predictor variables in a
multiple regression model are highly correlated with each other. This
can lead to unstable estimates of the regression coefficients and make
it difficult to determine the individual effect of each predictor on the
outcome variable.

```{r}
library (car)
# Calculate VIF
vif(model)
```

### **Diagnostic Plots for the MLR Model**

```{r, fig.height=8, fig.width=10}
# Generate diagnostic plots
par(mfrow =c(2 ,2))
plot(model)
par(mfrow =c(1 ,1))
```

### **Identify the Influential Observation as Indicared by Cook's Distance**

```{r}
# Calculate Cook's distance
cooks_distance <- cooks.distance(model)
# Identify influential observations
influential_obs <- which(cooks_distance > (4 / nrow(df)))
influential_obs
```

### **Remove Influential Observation and Refit the Model**

```{r}
# Remove influential observation
df_no_influential <- df[-influential_obs, ]
# Refit the model without influential observation
model_mlr_no_influential <- lm(CPI ~ ExchRate+LendIntRates, data = df_no_influential)
summary(model_mlr_no_influential)
```

### **Inferential Test for Normality and Homoscedasticity**

#### **Null and Alternative Hypotheses for Normality:**

- H0: The residuals of the model are normally distributed.
- H1: The residuals of the model are not normally distributed.

```{r}
# Normality test (Shapiro-Wilk test)
shapiro.test(residuals(model_mlr))
```

```{r}
# Normality test (Anderson-Darling test)
library(nortest)
ad.test(residuals(model_mlr))
```

```{r}
ggplot(data = data.frame(residuals = residuals(model_mlr)), aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot of Residuals", 
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles")
```

```{r}
ggplot(data = data.frame(residuals = residuals(model_mlr)), aes(x = residuals)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  labs(title = "Histogram of Residuals", 
       x = "Residuals", 
       y = "Frequency")
```

#### **Null and Alternative Hypotheses for Homoscedasticity:**

- H0: The residuals of the model have constant variance
  (homoscedasticity).
- H1: The residuals of the model do not have constant variance
  (heteroscedasticity).

```{r}
# Homoscedasticity test (Breusch-Pagan test)
library(lmtest)
bptest(model_mlr)
```

#### **Null and Alternative Hypotheses for Independence:**

- H0: The residuals of the model are independent (no autocorrelation).
- H1: The residuals of the model are not independent (autocorrelation
  present).

```{r}
# independence test (Durbin-Watson test)
dwtest(model_mlr)
```

## **Marginal Elasticities**

```{r}
model_elasticity <- lm(log(CPI) ~ log(ExchRate) + log(LendIntRates), data = df)
summary(model_elasticity)
```

### **The estimated equation is:**

$$\ln(\text{CPI})=-0.7043+1.5149\ln(\text{ExchRate})-0.5520\ln(\text{LendIntRates}).$$

Because **both the dependent and independent variables are in
logarithms**, this is a **log-log (double-log) regression model**. The
coefficients are interpreted as **elasticities**, **not multipliers**.

or equivalently,

$$CPI=e^{-0.7043}\times ExchRate^{1.5149}\times LendIntRates^{-0.5520}.$$

## **Interpretation of the coefficients**

### **1. Exchange Rate Elasticity**

The coefficient is

$$\beta_1=1.5149.$$

Interpretation:

Holding lending interest rates constant, a **1% increase in the exchange
rate** is associated with approximately a **1.515% increase in the
Consumer Price Index (CPI)**.

Since

$$1.515>1,$$

CPI is **elastic** with respect to the exchange rate.

This suggests that exchange rate depreciation has a
more-than-proportional effect on inflation.

### **2. Lending Interest Rate Elasticity**

The coefficient is

$$\beta_2=-0.5520.$$

Interpretation:

Holding exchange rates constant, a **1% increase in lending interest
rates** is associated with approximately a **0.552% decrease in CPI**.

This relationship is **inelastic** because

$$|{-0.552}|<1.$$

Economically, higher lending rates reduce borrowing and aggregate
demand, thereby lowering inflationary pressures.

## **Statistical significance**

### Exchange Rate

$$p<0.001$$

Highly significant.

There is very strong evidence that exchange rate movements affect CPI.

### **Lending Interest Rate**

$$p=0.000544$$

Also highly significant.

Higher lending rates significantly reduce inflation.

### **Goodness of fit**

#### **Multiple (R\^2)**

$$R^2=0.9348$$

#### **Interpretation:**

Approximately **93.48%** of the variation in CPI is explained jointly by
exchange rates and lending interest rates.

#### **Adjusted (R\^2)**

$$0.9291$$

Since it is very close to (R\^2), the model is not overfitting.

### **Overall model significance**

The F-statistic is

$$F=164.8$$

with

$$p=2.317\times10^{-14}.$$

#### **Interpretation:**

The model is highly statistically significant, indicating that the
explanatory variables jointly provide substantial explanatory power for
CPI.

#### **What can you compute from this model?**

Unlike a consumption model, **you cannot compute the Keynesian
multiplier**

$$k=\frac{1}{1-MPC},$$

because:

- there is **no consumption variable**,
- there is **no income variable**, and
- the coefficients are **elasticities**, not marginal propensities to
  consume.

Instead, you obtain:

- Exchange rate elasticity of inflation = **1.515**
- Lending interest rate elasticity of inflation = **−0.552**
- Overall explanatory power ((R\^2))
- Statistical significance of each determinant
- Predictive equation for CPI
- Evidence on exchange rate pass-through and the effect of monetary
  policy (through lending rates)

#### **A suitable interpretation for a research paper**

The estimated double-log regression model indicates that the exchange
rate has a positive and statistically significant effect on inflation.
Specifically, a 1% increase in the exchange rate is associated with an
estimated 1.515% increase in the Consumer Price Index (CPI), holding
lending interest rates constant. This suggests a highly elastic exchange
rate pass-through to domestic prices. Conversely, lending interest rates
have a negative and statistically significant effect on inflation, with
a 1% increase in lending rates associated with an estimated 0.552%
decrease in CPI. The model explains approximately 93.5% of the variation
in CPI ((R\^2 = 0.9348)), and the overall regression is highly
significant ((F = 164.8), (p \< 0.001)), indicating that exchange rates
and lending interest rates jointly provide a strong explanation of
inflation dynamics in the sample.

## **Demand Analysis**

```{r}
data <- read.csv("demand_data.csv")
head(data)
```

### **Estimate the Linear Model**

```{r}
model <- lm(log(Quantity) ~
              log(Price) +
              log(Income) +
              log(SubPrice) +
              log(CompPrice),
            data=data)

summary(model)
```

These are excellent results for illustrating **demand elasticity**.
Since you estimated a **double-log demand model**, every slope
coefficient is an elasticity. This makes the model ideal for teaching
and reporting economic concepts.

Your estimated model is

$$\ln(Q)=8.194-1.139\ln(P)+0.753\ln(Y)+0.537\ln(P_s)-0.407\ln(P_c)$$

where:

- 

  (Q) = Quantity demanded

- 

  (P) = Price of the good

- 

  (Y) = Consumer income

- (P_s) = Price of substitute goods

- (P_c) = Price of complementary goods

### **1. Price Elasticity of Demand**

The estimated own-price elasticity is

$$E_p=-1.139.$$

### **Interpretation**

Holding income, substitute prices, and complementary goods' prices
constant, a **1% increase in the price of the good decreases quantity
demanded by approximately 1.139%**.

Since

$$|E_p|=1.139>1,$$

the demand is **price elastic**. Consumers are relatively responsive to
price changes.

### **Economic implication**

A price increase is likely to reduce total revenue because the
percentage decline in quantity demanded exceeds the percentage increase
in price.

### **2. Income Elasticity of Demand**

The estimated income elasticity is

$$E_Y=0.753.$$

### **Interpretation**

Holding other variables constant, a **1% increase in consumer income
increases quantity demanded by approximately 0.753%**.

Since

$$0<E_Y<1,$$

the product is a **normal necessity**.

### **Economic implication**

Demand increases as consumers become wealthier, but at a slower rate
than income growth.

### **3. Cross-Price Elasticity (Substitute Goods)**

The estimated elasticity is

$$E_{Sub}=0.537.$$

### **Interpretation**

A **1% increase in the price of a substitute good increases demand for
this product by approximately 0.537%**, holding other variables
constant.

Because

$$E_{Sub}>0,$$

the goods are **substitutes**.

### **Examples**

- Coca-Cola and Pepsi
- Tea and Coffee
- Butter and Margarine

### **Economic implication**

If competitors raise their prices, consumers switch to your product.

### **4. Cross-Price Elasticity (Complementary Goods)**

The estimated elasticity is

$$E_{Comp}=-0.407.$$

### **Interpretation**

A **1% increase in the price of a complementary good decreases demand
for this product by approximately 0.407%**, holding other factors
constant.

Because

$$E_{Comp}<0,$$

the goods are **complements**.

### **Examples**

- Cars and fuel
- Printers and ink cartridges
- Bread and butter

### **Economic implication**

An increase in the price of a complementary product reduces demand for
the focal product.

#### **5. Statistical Significance**

All explanatory variables have p-values less than 0.001, indicating that
price, income, substitute prices, and complementary goods' prices all
have statistically significant effects on quantity demanded.

### **6. Model Fit**

The coefficient of determination is

$$R^2=0.9104.$$

### **Interpretation**

Approximately **91.04%** of the variation in quantity demanded is
explained by the four explanatory variables.

The adjusted (R\^2) of 0.9066 indicates that the model retains very
strong explanatory power after accounting for the number of predictors.

### **7. Overall Model Significance**

The overall regression is highly significant:

- (F = 241.3)
- (p \< 0.001)

This indicates that the explanatory variables jointly provide
substantial explanatory power for explaining demand.

### **Summary Table**

| Variable | Elasticity | Economic Interpretation | Classification |
|---------------|--------------:|---------------------------|---------------|
| Price | **−1.139** | 1% ↑ in price → 1.139% ↓ in demand | **Elastic demand** |
| Income | **0.753** | 1% ↑ in income → 0.753% ↑ in demand | **Normal necessity** |
| Substitute Price | **0.537** | 1% ↑ in substitute's price → 0.537% ↑ in demand | **Substitute goods** |
| Complement Price | **−0.407** | 1% ↑ in complement's price → 0.407% ↓ in demand | **Complementary goods** |

The model demonstrates several fundamental concepts in demand analysis
within a single regression:

- ✅ Demand function estimation
- ✅ Own-price elasticity of demand
- ✅ Income elasticity of demand
- ✅ Cross-price elasticity of demand (substitutes)
- ✅ Cross-price elasticity of demand (complements)
- ✅ Elastic vs. inelastic demand
- ✅ Normal vs. inferior goods (if income elasticity were negative)
- ✅ Statistical significance (t-tests and p-values)
- ✅ Overall model significance (F-test)
- ✅ Goodness of fit ((R\^2) and Adjusted (R\^2))

Because it captures nearly all of the core elasticity concepts in one
framework, this type of double-log demand model is widely used in
econometrics and is an excellent example for classroom demonstrations,
assignments, and applied research.

## **Consumption and Income Example**

```{r}
df2 <- read.csv("income.csv")
head(df2, 10)
```

## **Simple Linear Model**

### **Estimate the Impact of Income on Consumption**

```{r}
model2 <- lm(Consumption ~ Income, data = df2)
summary(model2)
```

### **Hypothesis for the Overall Significance of the Model (F-test)**

> Null hypothesis: The null model is better than the fitted model (the
> model does not explain a significant portion of the variance in
> consumption).

> Alternative hypothesis: The fitted model is better than the null model
> (the model explains a significant portion of the variance in
> consumption).

```{r}
library(report)
report(model2)
```

### **Estimate the Long-run Consumption Model**

```{r}
model3 <- lm(Consumption ~ Income-1, data = df2)
summary(model3)
```
